Efficiently Handling Guards in the Low Level Design of RPython's tracing JIT

### 2012-07-26 talk/vmil2012/paper.tex
consider some more operations as get/set, consider uint operations as numeric and print an overview of the operations considered "rest" at the end

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{wrapfig}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{relsize}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

@@

In this paper we describe and analyze how deoptimization works in the context
of tracing just-in-time compilers. What instructions are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

Although there are several publications about tracing just-in-time compilers, to
our knowledge, there are none that describe the use and implementation of
guards in this context. With the following contributions we aim to shed some
light (to much?) on this topic.
The contributions of this paper are:
\begin{itemize}
 \item
\end{itemize}

@@

The RPython language and the PyPy Project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language\bivab{mention speed}. The
Implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

@@

 * Tracing JITs
 * JIT Compiler
   * describe the tracing jit stuff in pypy
   * reference tracing the meta level paper for a high level description of what the JIT does
   * JIT Architecture
   * Explain the aspects of tracing and optimization

@@

Since tracing linearizes control flow by following one concrete execution,
not the full control flow of a program is observed.
The possible points of deviation from the trace are guard operations
that check whether the same assumptions as during tracing still hold.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to construct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

To do this reconstruction, it is necessary to take the values
of the SSA variables of the trace
and build interpreter stack frames.
Tracing aggressively inlines functions.
Therefore the reconstructed state of the interpreter
can consist of several interpreter frames.

@@

\begin{itemize}
    \item For small integer constants
        the payload contains the value of the constant.
    \item For other constants
        the payload contains an index into a per-loop list of constants.
    \item For SSA variables,
        the payload is the number of the variable.
    \item For virtuals,
        the payload is an index into a list of virtuals, see next section.
\end{itemize}

@@

% subsection Interaction With Optimization (end)

   * tracing and attaching bridges and throwing away resume data
   * compiling bridges
\bivab{mention that the failargs also go into the bridge}
% section Resume Data (end)

\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
% section Resume Data (end)
\section{Guards in the Backend}
\label{sec:Guards in the Backend}

@@

Once the bridge has been compiled the trampoline method stub is redirected to
the code of the bridge. In future if the guard fails again it jumps to the code
compiled for the bridge instead of bailing out. Once the guard has been
compiled and attached to the loop the guard becomes just a point where
control-flow can split. The loop after the guard and the bridge are just
conditional paths. \cfbolz{maybe add the unpatched and patched assembler of the trampoline as well?}

%* Low level handling of guards
%   * Fast guard checks v/s memory usage
%   * memory efficient encoding of low level resume data
%   * fast checks for guard conditions
%   * slow bail out
%
% section Guards in the Backend (end)

@@

The following analysis is based on a selection of benchmarks taken from the set
of benchmarks used to measure the performance of PyPy as can be seen
on\footnote{http://speed.pypy.org/}. The selection is based on the following
criteria \bivab{??}. The benchmarks were taken from the PyPy benchmarks
repository using revision
\texttt{ff7b35837d0f}\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}.
The benchmarks were run on a version of PyPy based on the
tag~\texttt{release-1.9} and patched to collect additional data about the
guards in the machine code
backends\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9}. All
benchmark data was collected on a MacBook Pro 64 bit running Max OS
10.7.4 \bivab{do we need more data for this kind of benchmarks} with the loop
unrolling optimization disabled\bivab{rationale?}.

@@

\bivab{should we rather count the trampolines as part of the guard data instead
of counting it as part of the instructions}

@@

LuaJIT, ...

@@

\section*{Acknowledgements}

\bibliographystyle{abbrv}
\bibliography{zotero,paper}

\end{document}

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{wrapfig}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{relsize}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage[colorinlistoftodos]{todonotes}

@@

\todo{add page numbers (draft) for review}
In this paper we describe and analyze how deoptimization works in the context
of tracing just-in-time compilers. What instructions are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

Although there are several publications about tracing just-in-time compilers, to
our knowledge, there are none that describe the use and implementation of
guards in this context. With the following contributions we aim to shed some
light (to much?) on this topic.
The contributions of this paper are:
\todo{more motivation}
\todo{extend}
\todo{contributions, description of PyPy's guard architecture, analysis on benchmarks}
\begin{itemize}
 \item
\end{itemize}

@@

The RPython language and the PyPy Project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language\bivab{mention speed}. The
implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

@@

 * Tracing JITs
 * Mention SSA
 * JIT Compiler
   * describe the tracing jit stuff in pypy
   * reference tracing the meta level paper for a high level description of what the JIT does
   * JIT Architecture
   * Explain the aspects of tracing and optimization

@@

Since tracing linearizes control flow by following one concrete execution,
not the full control flow of a program is observed.
The possible points of deviation from the trace are guard operations
that check whether the same assumptions observed during tracing still hold during execution.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to reconstruct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

To do this reconstruction, it is necessary to take the values of the SSA
variables of the trace and build interpreter stack frames.  Tracing
aggressively inlines functions, therefore the reconstructed state of the
interpreter can consist of several interpreter frames.

@@

\begin{itemize}
    \item For small integer constants
        the payload contains the value of the constant.
    \item For other constants
        the payload contains an index into a per-loop list of constants.
    \item For SSA variables,
        the payload is the number of the variable.
    \item For virtuals,
        the payload is an index into a list of virtuals, see next section.
\end{itemize}
\todo{figure showing linked resume-data}

@@

% subsection Interaction With Optimization (end)
\subsection{Compiling Side-Exits and Trace Stitching} % (fold)
\label{sub:Compiling side-exits and trace stitching}
   * tracing and attaching bridges and throwing away resume data
   * restoring the state of the tracer
     * keeping virtuals
   * compiling bridges
\todo{maybe mention that the failargs also go into the bridge}

% subsection Compiling side-exits and trace stitching (end)
% section Resume Data (end)

\todo{set line numbers to the line numbers of the rpython example}
\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
% section Resume Data (end)
\section{Guards in the Backend}
\label{sec:Guards in the Backend}

@@

Once the bridge has been compiled the trampoline method stub is redirected to
the code of the bridge. In future if the guard fails again it jumps to the code
compiled for the bridge instead of bailing out. Once the guard has been
compiled and attached to the loop the guard becomes just a point where
control-flow can split. The loop after the guard and the bridge are just
conditional paths. \todo{add figure of trace with trampoline and patched guard to a bridge}
%* Low level handling of guards
%   * Fast guard checks v/s memory usage
%   * memory efficient encoding of low level resume data
%   * fast checks for guard conditions
%   * slow bail out
%
% section Guards in the Backend (end)

@@

The following analysis is based on a selection of benchmarks taken from the set
of benchmarks used to measure the performance of PyPy as can be seen
on.\footnote{http://speed.pypy.org/} The benchmarks were taken from the PyPy benchmarks
repository using revision
\texttt{ff7b35837d0f}.\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}
The benchmarks were run on a version of PyPy based on the
tag~\texttt{release-1.9} and patched to collect additional data about the
guards in the machine code
backends.\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9} All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.7.4 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

@@

\todo{resume data size estimates on 64bit}
\todo{figure about failure counts of guards (histogram?)}
\todo{integrate high level resume data size into Figure \ref{fig:backend_data}}
\todo{count number of guards with bridges for \ref{fig:ops_count}}
\todo{add resume data sizes without sharing}

@@

\todo{look into tracing papers for information about guards and deoptimization}
LuaJIT \todo{link to mailing list discussion}

@@

\todo{conclusion}

\section*{Acknowledgements}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
\listoftodos
\end{document}
